For Induction-Induction, Induction is Enough

### 2019-11-11 paper/universaliit.tex
more small fixes, now looking at related work

\begin{enumerate}
\item Definition of untyped syntax (as a family of inductive datatypes) together
  with typing judgments (as inductive relations on the untyped syntax), and
  construction of a model of the theory of IIT signatures from well-formed
  terms, denoted $S$.
\item Construction of a morphism $\rec:S\to M$ for arbitrary $M$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given model. The idea is that given a syntactic context $\Gamma$ and a
    semantic context $\model{\Gamma}$ of the model $M$, we have
    $\Gamma \sim \model{\Gamma}$ if and only if $\rec\,\,\Gamma
    = \model{\Gamma}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\Gamma}$ of
    % the model $M$ if and only if $\Gamma$ relates to $\rec\,{\Gamma}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:S\to M$ satisfies the relation. For example, for any syntactic context
    $\Gamma$ we have $\Gamma\sim f\,\Gamma$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.
\subsection{Syntactic Model}

@@

\subsubsection{Untyped syntax}
The untyped syntax is defined as the following inductive datatype. Variables are
modeled as de Bruijn indices, i.e.\ as natural numbers pointing to a position in
the context. We use the additional default constructor $\erru:\Tmu$ in case of
error (ill-scoped substitution). The typing judgments will not mention $\erru$.
\todo{András: prettify below maybe}
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4-5) Metatheoretic parameters} \\[0.5em]
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(6) Default value} \\[0.5em]
  & \erru && : \Tmu
\end{alignat*}

@@

Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
\,(A[\sigma])\,(B[\varu\, 0 \,\consu\, \wkO\,\sigma])$, and thus we need to define
$\wkO$, the weakening of substitutions.  The basic idea is to increment the de
Bruijn indices of all the variables.  Actually this is not so simple because of
the $\Piu$ type: indeed, we want to define $\wkO\,(\Piu\,A\,B)$ as the $\Pi$ type
of the weakening of $A$ and $B$, but here, $B$ must be weakened with respect to
the second last variable of the context, rather than the last one.  For this
reason, we need to generalize the weakening as occuring anywhere in the context.
\begin{alignat*}{5}
  & \wk_n && :  \Tyu\ra\Tyu \\
  & \wk_n && :  \Tmu\ra\Tmu \\
  % & \wk && : \N \ra \N\ra\N \\
  & \wkO && : \Subu\ra \Subu
  \end{alignat*}
  The natural number $n$ specifies at which position of the context the
  weakening occurs.
  Here, $\wkO$ weakens with respect to the last variable.

@@

  However, to construct the initial model of the QIIT, we need to define
  the full substitution calculus:
\begin{alignat*}{5}
  & \blank[\blank] && : \Tyu\ra\Subu\ra\Tyu \\
  & \blank[\blank] && : \Tmu\ra\Subu\ra\Tmu \\
  & \blank \circ \blank && : \Subu\ra\Subu\ra \Subu
  \end{alignat*}
These can be defined either by iterating unary substitutions, or by recursion
on untyped syntax: the two ways yield provably equal definitions. In the
following, we assume that it is defined by recursion. We also make use of the
following definition:
  \begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \keep && : \Subu \ra \Subu \\
   &&& := \lambda \sigma. (\varu\, 0 \,\consu\, \wkO\,\sigma)
  \end{alignat*}
  The idea is that if $\sigma$ is a substitution between contexts $\Gamma$ and
  $\Delta$, then $\keep\,\sigma$ is a substitution between contexts $\Gamma\ext A[\sigma]$
  and $\Delta\ext A$ for any type $A$. This occurs when defining
  $(\Piu\,A\,B)[\sigma]$ as $\Piu\,(A[\sigma])\,(B[\keep \,\sigma])$.

@@

  \subsubsection{Exchange laws: weakening and substitution}
  Many lemmas for types and terms are shown by induction on the untyped syntax.
  Below, $Z$ denotes either a term or a type.
\begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \wk\mhyphen\wk && : \wk_{n+p+1} (\wk_n\,Z) = \wk_n(\wk_{n+p}\,Z)\\
  % n-subZ-wkZ : ∀ n A z → (lifZZ n A) [ n ↦ z ]Z  ≡ A
  & \wk_n[n] && : (\wk_{n} Z)[ n := z] = Z \\
  % lifZ-l-subV : ∀ n p u x → lifZZ (n + p) (x [ n ↦ u ]V) ≡ (lifZV (S (n + p)) x) [ n ↦ (lifZZ p u) ]V
  & \wk_+[] && : (\wk_{n+p+1} Z)[ n := \wk_p u] = \wk_{n+p}(Z[n := u]) \\
  % l-subV-lifZV : ∀ Δ u n x → (lifZV n x) [ (S (n + Δ)) ↦ u ]V  ≡ lifZZ n (x [ (n + Δ) ↦ u ]V)
  & \wk[+] && : (\wk_{n} Z)[ n+p+1 :=  u] = \wk_{n}(Z[n+p := u]) \\
  % l-subV-l-subV : ∀ n p z u x →  (x [ n ↦ u ]V) [ (n + p) ↦ z ]Z  ≡ (x [ (S (n + p)) ↦ z ]V) [ n ↦ (u [ p ↦ z ]Z) ]Z
  & [][+] && : Z[n :=u][n+p:=z] = Z[n+p+1:=z][n:= (u[p:=z])]
  \end{alignat*}
  Below, $\keep^n$ denotes $\keep$ iterated $n$ times.
\begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  % liftV=wkS
  & [\keep^n\mhyphen \wkO] && : Z[\keep^n (\wkO\,\sigma)] = \wk_n (Z[\keep^n\,\sigma])
  \\
  % lifZₛV : ∀ n xp σp Zp → (lifZV n xp [ iZer n keep (Zp ∷ σp) ]V) ≡ (xp [ iZer n keep σp ]V)
  & \wk_n[\keep^n\mhyphen\cons]  & &
  :
  (\wk_n Z) [\keep^n (u \consu \sigma)] = Z [ \keep^n \sigma]
  \\
  % l-sub[]V : ∀ n x z σ
    % ((x [ n := z ]V) [ iZer n keep σ ]Z) = (x [ iZer (S n) keep σ ]V) [ n := (z [ σ ]Z) ]Z
   & [:=][\keep] && : Z [ n:= u][\keep^n \sigma] =  Z[\keep^{n+1}\,\sigma][n := u[\sigma]]
  \end{alignat*}
As particular cases for $n=0$, we get
\begin{alignat*}{5}
  & {}{\circ}{}\wkO & & : \sigma \circ (\wkO \tau) = \wkO (\sigma \circ \tau) \\
  & \wkO{}{\circ}{}, && : \wkO\,\sigma \circ (t \consu \tau ) = \sigma \circ \tau \\
  & [\wkO] && : t[\wkO\,\sigma] = \wk_0 (t[\sigma]) \\
  & \wk_0[\cons] && : (\wk_0 Z)[u \consu \sigma] = Z[\sigma] \\
  & [0:=][] && : Z [ 0:= u][ \sigma] =  Z[\keep\,\sigma][0 := u[\sigma]]
\end{alignat*}
Finally, we show the following:
\begin{alignat*}{5}
  & [][] && : Z[\sigma][\tau] = Z[\sigma\circ\tau]
  \\
  & \ass && : (\sigma\circ \delta)\circ\tau = \sigma\circ (\delta\circ\tau)
  \end{alignat*}
We defer laws for identity substitutions after the definition of the typing
judgments, as the proofs require that some inputs are well-typed.

@@

  \subsubsection{Typing judgments}
  \label{sec:typing_judgments}
  The typing judgments are defined as the following inductive datatype indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \Gamma\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \Gamma\vdash )  \ra  (\Gamma\vdash A)  \ra
  \Gamma\extu A \vdash \\
  & \consw && :
    (\Delta \vdash) \ra
    (\Gamma\vdash \sigma \Rightarrow\Delta)\ra
    ( \Delta \vdash A) \ra
    ( \Gamma \vdash t \in A[\sigma]) \ra
    \Gamma \vdash t \consu \sigma \Rightarrow \Delta\extu A
   \\
  & \varw  && : (\Gamma \vdash n \invar A) \ra \Gamma \vdash \varu n \in A \\
  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash 0 \invar \wku \, A \\
  & \vsw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra(\Gamma \vdash n \invar A) \ra (\Gamma \vdash B) \ra \Gamma \extu  B
  \vdash S\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\Gamma \vdash)\ra \Gamma \vdash \Uu \\
  & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Piw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
  \Elu\,a \vdash B)\ra\Gamma \vdash \Piu \, a \, B \\
  & \appw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Piu \, a \, B )
    \ra
    (\Gamma \vdash u \in \Elu\,a)
    \ra
    \Gamma \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Inductive parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \Gamma \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
\end{alignat*}
There is possibility of redundancy in the arguments of the constructors. Here,
we are ``{paranoid}'', so that we get more inductive hypotheses when performing
recursion.

@@

Up until the construction of the syntactic model, UIP is not used. Function
extensionality on the other hand is necessary because the untyped metatheoretic
$\Pi$ takes a metatheoretic function as an argument. An example induction step
that uses it is in the type preservation proof for identity substitutions (this
is an equation satisfied by a model of the QIIT), in particular in the case
$(\Pim\,T\, A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is
equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction
hypothesis states that $(t:T)\ra (A\,t)[\id]=A\,t$.

@@

This approach is an alternative version of Streicher's method for interpreting
preterms in an arbitrary model\cite{streichersemantics}. Streicher first defines
a family of partial maps from the presyntax to a model, then shows that the maps
are total on well-formed input. We have found that our approach is significantly
easier to formalise. Too see why, note that the right notion of partial map in
type theory, which does not presume decidable definedness, is fairly
heavyweight:
\[
  \mathsf{PartialMap}\,A\,B := A\ra (P : \Set)\times\isaprop\,P\times (P \ra B)
\]
In the above definition, we notice an opportunity for converting a fibered
definition of a type family into an indexed one; if we drop the propositionality
for $P$ for the time being, we may equivalently return a family indexed over
$B$, which is exactly just a relation $A\ra B\ra\Set$. Then, in our approach, we
recover uniqueness of $P$ through the functionality requirement on the $A\ra
B\ra\Set$ relation, and totality by already assuming well-formedness of $A$. In
type theory, using indexed families instead of display maps is a common
convenience, since the former are natively supported in type theory, while the
latter require carrying around auxiliary propositional equalities.

@@

Given a model $M$ of the QIIT, we define the functional relation satisfied by
the initial morphism $\rec : S\to M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\Gamma$ is a context in $S$ and $\model{\Gamma}$ is a semantic context
(i.e.\ a context of the model $M$), we want to define a type
$\Gamma\sim \model{\Gamma}$ equivalent to $\rec\,\,\Gamma=\model{\Gamma}$. Of
course, at this stage, $\rec$ is not available yet since the point of
defining this relation is to construct $\rec$ in the end.

For a type $A$ in a context $\Gamma$, we want to define a relation
$A\sim \model{A}$ that is equivalent to $\rec\,\,A=\model{A}$.
% where $\model{A}$ is a
% semantic type in a semantic context $\model{\Gamma}$.
For this equality to make sense, the semantic type $\model{A}$ must live
in the semantic context $\rec\,\,\Gamma$. But again, $\rec$ is not
yet available at this stage. Exploiting the expected equivalence between
$\Gamma\sim\model\Gamma$ and $\rec\,\,\Gamma=\model\Gamma$, we may consider defining $A\sim\model{A}$ under the
hypotheses that $\model{A}$ lies in a semantics context $\model\Gamma$ which is
related to $\Gamma$. Then, the type of the relation for types is
\[
  (\Gamma : \syn\Con)\ra (A : \syn\Ty\syn\Gamma) \ra
  (\model\Gamma : \model\Con)
  \ra
  (\Gamma \sim \model\Gamma)
  \ra
  (\model{A}:\model\Ty\model\Gamma)
  \ra
  \Set
\]
Note that the relation on contexts
must be defined mutually with the relation on types (see for example the case of
context extension), but here, the relation on contexts appears as the type of an
argument of the relation on types.
% We leave this possibility aside,
Our metatheory does not support this kind of recursive-recursive
definitions\footnote{And to our knowledge there is no established semantics for
it in existing literature.}, so we instead just remove the hypothesis
$\Gamma\sim\model\Gamma$ from the list of arguments.  We proceed similarly for
terms and substitutions. Actually, this removal is not without harm. For
example, consider relating the empty substitution
$\Gamma\vdash \epsilonu \Rightarrow \emptyu $ to a semantic substitution
$\model\sigma : \model\Sub\,\model\Gamma\,\model\Delta$. We would like to assert
that $\model\sigma$ equals the empty semantic substitution $\model\epsilon$, but
this is not possible because typechecking requires that $\model\Delta$ is the
empty semantic context. This is precisely what ensured the removed hypothesis
$\syn\cdot \sim \model\Delta$.  Our way out here is to state that $\model\sigma$
is related to the empty substitution if the target semantic context $\model\Delta$
is empty, and, acknowledging this equality, if $\model\sigma$ is the empty
substitution.
% The idea is that we will first prove that there is atmost one semantic type
% which relates to the syntactic one, and then later we will provide such a
% semantic type under the hypothesis that the contexts are related.

>>>>

\begin{enumerate}
\item Definition of untyped syntax (as a family of inductive datatypes) together
  with typing judgments (as inductive relations on the untyped syntax), and
  construction of a model of the theory of IIT signatures from well-formed
  terms, denoted $\mathsf{S}$.
\item Construction of a morphism $\rec:\mathsf{S}\to M$ for arbitrary $M$, by:
  \begin{enumerate}
  \item defining a relation $\blank \sim \blank$ between the well-typed syntax
    and a given model. The idea is that given a syntactic context $\Gamma$ and a
    semantic context $\model{\Gamma}$ of the model $M$, we have
    $\Gamma \sim \model{\Gamma}$ if and only if $\rec\,\,\Gamma
    = \model{\Gamma}$, and similarly for types, terms, and substitutions;
    % semantic context $\model{\Gamma}$ of
    % the model $M$ if and only if $\Gamma$ relates to $\rec\,{\Gamma}$;
    % enjoyed by the initial morphism from the
    % syntax to a given model;
  \item showing that this relation is functional.
    \end{enumerate}
  \item Proving the uniqueness of this morphism by showing that any morphism
    $f:\mathsf{S}\to M$ satisfies the relation. For example, for any syntactic context
    $\Gamma$ we have $\Gamma\sim f\,\Gamma$.
      % TODO donner composantes sur les contextes
\end{enumerate}
The next sections detail each of these steps.
\subsection{Syntactic Model}

@@

\subsubsection{Untyped syntax}
The untyped syntax is defined as the following inductive datatype. Variables are
modeled as de Bruijn indices, i.e.\ as natural numbers pointing to a position in
the context. We use the additional default constructor $\erru:\Tmu$ in case of
error (ill-scoped substitution). The typing judgments will not mention $\erru$.
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \Conu && : \Set \\
  & \Tyu  && : \Set \\
  & \Subu  && : \Set \\
  % & \Subpre  && : \Set \\
  & \Tmu  && : \Set \\
  & \emptyu && : \Conu \\
    & \epsilonu && : \Subu \\
  & \blank\extu\blank && : \Conu\ra\Tyu\ra\Conu \\
  & \blank\consu\blank && : \Subu\ra\Tmu\ra\Subu \\
  & \varu  && : \N \ra \Tmu \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uu && : \Tyu \\
  & \Elu && : \Tmu\ra\Tyu \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Piu && : \Tmu\ra\Tyu\ra\Tyu \\
  & \blank\appu\blank && : \Tmu\ra\Tmu\ra \Tmu \\
  & \rlap{(4-5) Metatheoretic parameters} \\[0.5em]
  & \Pimu && : (T:\Set)\ra(T\ra\Tyu)\ra\Tyu \\
  & \Piiu && : (T:\Set)\ra(T\ra\Tmu)\ra\Tmu \\
  & \blank\appmu \blank && : \Tmu\ra(\alpha:T) \ra\Tmu\\
  & \rlap{(6) Default value} \\[0.5em]
  & \erru && : \Tmu
\end{alignat*}

@@

Note that $(\Piu \,A\,B)[\sigma]$ should be defined as $\Piu
\,(A[\sigma])\,(B[\varu\, \mathsf{0} \,\consu\, \wkO\,\sigma])$, and thus we need to define
$\wkO$, the weakening of substitutions.  The basic idea is to increment the de
Bruijn indices of all the variables.  Actually this is not so simple because of
the $\Piu$ type: indeed, we want to define $\wkO\,(\Piu\,A\,B)$ as the $\Pi$ type
of the weakening of $A$ and $B$, but here, $B$ must be weakened with respect to
the second last variable of the context, rather than the last one.  For this
reason, we need to generalize the weakening as occuring anywhere in the context.
\begin{alignat*}{5}
  & \wk_n && :  \Tyu\ra\Tyu \\
  & \wk_n && :  \Tmu\ra\Tmu \\
  % & \wk && : \N \ra \N\ra\N \\
  & \wkO && : \Subu\ra \Subu
  \end{alignat*}
  The natural number $n$ specifies at which position of the context the
  weakening occurs.
  Here, $\wkO$ weakens with respect to the last variable.

@@

  However, to construct the initial model of the QIIT, we need to define
  the full substitution calculus:
\begin{alignat*}{5}
  & \blank[\blank] && : \Tyu\ra\Subu\ra\Tyu \\
  & \blank[\blank] && : \Tmu\ra\Subu\ra\Tmu \\
  & \blank \circ \blank && : \Subu\ra\Subu\ra \Subu
  \end{alignat*}
These can be defined either by iterating unary substitutions, or by recursion
on untyped syntax: the two ways yield provably equal definitions. In the
following, we assume that it is defined by recursion. We also make use of the
following definition:
  \begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \keep && : \Subu \ra \Subu \\
   &&& := \lambda \sigma. (\varu\, \mathsf{0} \,\consu\, \wkO\,\sigma)
  \end{alignat*}
  The idea is that if $\sigma$ is a substitution between contexts $\Gamma$ and
  $\Delta$, then $\keep\,\sigma$ is a substitution between contexts $\Gamma\ext A[\sigma]$
  and $\Delta\ext A$ for any type $A$. This occurs when defining
  $(\Piu\,A\,B)[\sigma]$ as $\Piu\,(A[\sigma])\,(B[\keep \,\sigma])$.

@@

  \subsubsection{Exchange laws: weakening and substitution}
  Many lemmas for types and terms are shown by induction on the untyped syntax.
  Below, $Z$ denotes either a term or a type.
\begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  & \wk\mhyphen\wk && : \wk_{n+p+1} (\wk_n\,Z) = \wk_n(\wk_{n+p}\,Z)\\
  % n-subZ-wkZ : ∀ n A z → (lifZZ n A) [ n ↦ z ]Z  ≡ A
  & \wk_n[n] && : (\wk_{n}\,Z)[ n := z] = Z \\
  % lifZ-l-subV : ∀ n p u x → lifZZ (n + p) (x [ n ↦ u ]V) ≡ (lifZV (S (n + p)) x) [ n ↦ (lifZZ p u) ]V
  & \wk_+[] && : (\wk_{n+p+1}\,Z)[ n := \wk_p\,u] = \wk_{n+p}\,(Z[n := u]) \\
  % l-subV-lifZV : ∀ Δ u n x → (lifZV n x) [ (S (n + Δ)) ↦ u ]V  ≡ lifZZ n (x [ (n + Δ) ↦ u ]V)
  & \wk[+] && : (\wk_{n}\,Z)[ n+p+1 :=  u] = \wk_{n}\,(Z[n+p := u]) \\
  % l-subV-l-subV : ∀ n p z u x →  (x [ n ↦ u ]V) [ (n + p) ↦ z ]Z  ≡ (x [ (S (n + p)) ↦ z ]V) [ n ↦ (u [ p ↦ z ]Z) ]Z
  & [][+] && : Z[n :=u][n+p:=z] = Z[n+p+1:=z][n:= (u[p:=z])]
  \end{alignat*}
  Below, $\keep^n$ denotes $\keep$ iterated $n$ times.
\begin{alignat*}{5}
  % liftV (S (n + p)) (liftV n q) ≡ liftV n (liftV (n + p) q)
  % liftV=wkS
  & [\keep^n\mhyphen \wkO] && : Z[\keep^n\,(\wkO\,\sigma)] = \wk_n (Z[\keep^n\,\sigma])
  \\
  % lifZₛV : ∀ n xp σp Zp → (lifZV n xp [ iZer n keep (Zp ∷ σp) ]V) ≡ (xp [ iZer n keep σp ]V)
  & \wk_n[\keep^n\mhyphen\cons]  & &
  :
  (\wk_n\,Z) [\keep^n\, (u \consu \sigma)] = Z [ \keep^n \sigma]
  \\
  % l-sub[]V : ∀ n x z σ
    % ((x [ n := z ]V) [ iZer n keep σ ]Z) = (x [ iZer (S n) keep σ ]V) [ n := (z [ σ ]Z) ]Z
   & [:=][\keep] && : Z [ n:= u][\keep^n\,\sigma] =  Z[\keep^{n+1}\,\sigma][n := u[\sigma]]
  \end{alignat*}
As particular cases for $n=0$, we get
\begin{alignat*}{5}
  & {}{\circ}{}\wkO & & : \sigma \circ (\wkO \tau) = \wkO (\sigma \circ \tau) \\
  & \wkO{}{\circ}{}, && : \wkO\,\sigma \circ (t \consu \tau ) = \sigma \circ \tau \\
  & [\wkO] && : t[\wkO\,\sigma] = \wk_0 (t[\sigma]) \\
  & \wk_0[\cons] && : (\wk_0 Z)[u \consu \sigma] = Z[\sigma] \\
  & [0:=][] && : Z [ 0:= u][ \sigma] =  Z[\keep\,\sigma][0 := u[\sigma]]
\end{alignat*}
Finally, we show the following:
\begin{alignat*}{5}
  & [][] && : Z[\sigma][\tau] = Z[\sigma\circ\tau]
  \\
  & \ass && : (\sigma\circ \delta)\circ\tau = \sigma\circ (\delta\circ\tau)
  \end{alignat*}
We defer laws for identity substitutions after the definition of the typing
judgments, as the proofs require that some inputs are well-typed.

@@

  \subsubsection{Typing judgments}
  \label{sec:typing_judgments}
  The typing judgments are defined as the following inductive datatype indexed over the
  untyped syntax:
\begin{alignat*}{5}
  & \rlap{(1) Substitution calculus} \\[0.5em]
  & \blank \vdash && : \Conu\ra \Set \\
  & \blank \vdash \blank  && : \Conu\ra \Tyu\ra\Set \\
  & \blank \vdash \blank \invar \blank  && : \Conu\ra \N \ra \Tyu\ra \Set \\
  & \blank \vdash \blank \in \blank  && : \Conu\ra \Tmu\ra \Tyu\ra \Set \\
  & \blank \vdash \blank \Rightarrow \blank  && : \Conu\ra \Subu \ra \Conu\ra \Set \\
  % & \Subpre  && : \Set \\
  & \emptyw && : \emptyu \vdash \\
  & \epsilonw && : \Gamma\vdash \epsilonu \Rightarrow \emptyu \\
  & \blank\extw\blank && : ( \Gamma\vdash )  \ra  (\Gamma\vdash A)  \ra
  \Gamma\extu A \vdash \\
  & \consw && :
    (\Delta \vdash) \ra
    (\Gamma\vdash \sigma \Rightarrow\Delta)\ra
    ( \Delta \vdash A) \ra
    ( \Gamma \vdash t \in A[\sigma]) \ra
    \Gamma \vdash t \consu \sigma \Rightarrow \Delta\extu A
   \\
  & \varw  && : (\Gamma \vdash n \invar A) \ra \Gamma \vdash \varu n \in A \\
  & \vzw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra\Gamma \extu  A \vdash \mathsf{0} \invar \wku \, A \\
  & \vsw  && : (\Gamma\vdash)\ra(\Gamma\vdash A)\ra(\Gamma \vdash n \invar A) \ra (\Gamma \vdash B) \ra \Gamma \extu  B
  \vdash \mathsf{S}\,n \invar \wku \, A \\
  & \rlap{(2) Universe} \\[0.5em]
  & \Uw && : (\Gamma \vdash)\ra \Gamma \vdash \Uu \\
  & \Elw && : (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra\Gamma \vdash \Elu\,a \\
  & \rlap{(3) Inductive parameters} \\[0.5em]
  & \Piw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
  \Elu\,a \vdash B)\ra\Gamma \vdash \Piu \, a \, B \\
  & \appw && :
    (\Gamma \vdash)\ra (\Gamma \vdash a \in \Uu)\ra(\Gamma \extu
    \Elu\,a \vdash B)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Piu \, a \, B )
    \ra
    (\Gamma \vdash u \in \Elu\,a)
    \ra
    \Gamma \vdash t \appu u \in  B [0 := u] \\
  & \rlap{(4) Inductive parameters} \\[0.5em]
  & \Pimw && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t) \ra
    \Gamma \vdash \Pimu \,T\,A
    \\
    & \appmw  && :
    (T:\Set)\ra(A : T\ra \Tyu)\ra(\Gamma \vdash)\ra
    ((t : T)\ra\Gamma\vdash A\,t)
    \\
    & &&
    \ra(\Gamma \vdash t \in \Pimu \,T\,A)
    \ra (u : T) \ra \Gamma \vdash t\,\appmu\,u \in A\,u
\end{alignat*}
There is possibility of redundancy in the arguments of the constructors. Here,
we are ``{paranoid}'', so that we get more inductive hypotheses when performing
recursion.

@@

Up until the construction of the syntactic model, UIP is not used. Function
extensionality on the other hand is necessary because the untyped metatheoretic
$\Pi$ takes a metatheoretic function as an argument. An example induction step
that uses function extensionality is in the type preservation proof for identity substitutions (this
is an equation satisfied by a model of the QIIT), in particular in the case
$(\Pim\,T\, A)[\id]=\Pim\,T\,A$. Indeed, the left hand side of this equation is
equal to $\Pim\,T\,(\lambda t.(A\,t)[\id])$ by definition, whereas the induction
hypothesis states that $(t:T)\ra (A\,t)[\id]=A\,t$.

@@

This approach is an alternative version of Streicher's method for interpreting
preterms in an arbitrary model \cite{streichersemantics}. Streicher first defines
a family of partial maps from the presyntax to a model, then shows that the maps
are total on well-formed input. We have found that our approach is significantly
easier to formalise. Too see why, note that the right notion of partial map in
type theory, which does not presume decidable definedness, is fairly
heavyweight:
\[
  \mathsf{PartialMap}\,A\,B := A\ra \big((P : \Set)\times\isaprop\,P\times (P \ra B)\big)
\]
In the above definition, we notice an opportunity for converting a fibered
definition of a type family into an indexed one; if we drop the propositionality
for $P$ for the time being, we may equivalently return a family indexed over
$B$, which is exactly just a relation $A\ra B\ra\Set$. Then, in our approach, we
recover uniqueness of $P$ through the functionality requirement on the $A\ra
B\ra\Set$ relation, and totality by already assuming well-formedness of $A$. In
type theory, using indexed families instead of display maps is a common
convenience, since the former are natively supported, while the
latter require carrying around auxiliary propositional equalities.

@@

Given a model $M$ of the QIIT, we define the functional relation satisfied by
the initial morphism $\rec : \mathsf{S}\to M$ by recursion on the typing judgments.
% As these judgments are propositions, we will sometimes
If $\Gamma$ is a context in $S$ and $\model{\Gamma}$ is a semantic context
(i.e.\ a context of the model $M$), we want to define a type
$\Gamma\sim \model{\Gamma}$ equivalent to $\rec\,\,\Gamma=\model{\Gamma}$. Of
course, at this stage, $\rec$ is not available yet since the point of
defining this relation is to construct $\rec$ in the end.

For a type $A$ in a context $\Gamma$, we want to define a relation
$A\sim \model{A}$ that is equivalent to $\rec\,\,A=\model{A}$.
% where $\model{A}$ is a
% semantic type in a semantic context $\model{\Gamma}$.
For this equality to make sense, the semantic type $\model{A}$ must live
in the semantic context $\rec\,\,\Gamma$. But again, $\rec$ is not
yet available at this stage. Exploiting the expected equivalence between
$\Gamma\sim\model\Gamma$ and $\rec\,\,\Gamma=\model\Gamma$, we may consider defining $A\sim\model{A}$ under the
hypotheses that $\model{A}$ lies in a semantics context $\model\Gamma$ which is
related to $\Gamma$. Then, the type of the relation for types is
\[
  (\Gamma : \syn\Con)\ra (A : \syn\Ty\,\syn\Gamma) \ra
  (\model\Gamma : \model\Con)
  \ra
  (\Gamma \sim \model\Gamma)
  \ra
  (\model{A}:\model\Ty\,\model\Gamma)
  \ra
  \Set
\]
Note that the relation on contexts
must be defined mutually with the relation on types (see for example the case of
context extension), but here, the relation on contexts appears as the type of an
argument of the relation on types.
% We leave this possibility aside,
We want to avoid using such recursive-recursive
definitions as they are not allowed by the elimination principles of indexed inductive types, so we instead just remove the hypothesis
$\Gamma\sim\model\Gamma$ from the list of arguments.  We proceed similarly for
terms and substitutions. Actually, this removal is not without harm. For
example, consider relating the empty substitution
$\Gamma\vdash \epsilonu \Rightarrow \emptyu $ to a semantic substitution
$\model\sigma : \model\Sub\,\model\Gamma\,\model\Delta$. We would like to assert
that $\model\sigma$ equals the empty semantic substitution $\model\epsilon$, but
this is not possible because typechecking requires that $\model\Delta$ is the
empty semantic context. This is precisely what was ensured by the hypothesis
$\syn\cdot \sim \model\Delta$ we removed.  Our way out here is to state that $\model\sigma$
is related to the empty substitution if the target semantic context $\model\Delta$
is empty, and, acknowledging this equality, if $\model\sigma$ is the empty
substitution.
% The idea is that we will first prove that there is atmost one semantic type
% which relates to the syntactic one, and then later we will provide such a
% semantic type under the hypothesis that the contexts are related.
